phRAIDER: Pattern-Hunter based Rapid Ad Initio \\ Identification of Elemenrary Repeats

### 2015-09-19 manuscript/JCB/manuscript.tex

We can now modif the previous definition of elementary repeats as
follows:
\begin{definition} Given a genomic sequence $G$, an integer $f$, and a
  spaced seed $s$, a sequence descriptor $r$ describes an elementary
  repeat if it meets the four (moified) requirements of an elementary repeat:
\begin{itemize}
\item \underline{Structure requriement}: $s$ covers $r$.
\item \underline{Frequency requirement}: There are at least $f$
  substrings of $G$ that match $r$.
\item \underline{Minimality requirement:} For every string $t$ in the
  decomposition of $r$ w.r.t $s$, the number of occurences of $t$ in
  $G$ is equal to the number of occurneces of $r$ in the genome.
\item \underline{Maximality requirement:} There is no sequence
  descriptor $r'$ of $r$ that contains $r$ as a proper substring and
  satisfies conditions 1-1.
\end{itemize}
\end{definition}

@@

\begin{algorithm}
\caption{This is the algorthm caption.}
\begin{algorithmic}
  \Function{sameOffset}{$l$-mer $v_1$, $l$-mer $v_2$}
    \comment{Is the distance between the first two occurences of $v$ equal to the}
    \comment{the distance betwen the most recent two occurences?}
    \State \Return $H[v_2][0] - H[v_1][0]  == H[v_2][-1] - H[v_1][-1]$
  \EndFunction
  \State
  \Function{phRAIDER}{Genome $G$, Seed $s$, MinFrequence $f$}
    \comment{We are going to maintain a $Q$ of up to $|s|$ families, representing the families}
    \comment{of the last $|L|$ $l$-mers seen, sorted by the position of those last $l$-mers.}
    \State Queue$<$Family$>$ $Q$;
    \State
    \comment{Iterate over all $l$-mer positions in $G$}
    \For {$i \leftarrow 1 \to |G|-|s|$}

       \comment{Remove any families whose last $l$-mer did not occur in the last $i$ bases.}
       \While{$i - $location($Q$, finish, last, end)$ > |s|$}
         \State $Q.$dequeue()
       \EndWhile
       \State
       \comment{Get the $l$-mer occuring at $G_i$ and remove letters corresponding to $0$s in $s$.}
       \State $v \leftarrow \mbox{seeded}(G[i],x)$ 
       \State
       \comment{Add to list of $v$'s positions.}
       \State $H[v].push(i)$
       \State
       \If {$|H|(v)| == 2$}
          \comment{$v$ either is the beginning of a family, or should be combined with a}
          \comment{previous $l$-mer to form a family.}
          \State $F = \mbox{arg\_find}_{F \in Q}\{\mbox{sameOffset}(F.\mbox{start}(), v)\}$
          \If {$F$}
             \State $Q.$requeue($F$)
          \EndIf
          \State last($Q$).addLmer($v$)   
       \ElsIf {$|H[v]| > 2$}
          \comment{Is $v$ the proper continuation of a family, or does it need to be split off?}
          \State $F \leftarrow v.$family()
          \If {$H[F.\mbox{lastSeen}(i)][-1] < H[Q.\mbox{front}().\mbox{lastSeen}()]$}
              \comment{$v$ belongs to a family whose last instance is too far back.}
              \If {$v \neq F.$first$()$}
                  \comment{$v$ is not the first $l$-mer of its family}
                  \State $F' \leftarrow F.$split($F.v$)
              \Else
                  \State $F' \leftarrow F.$split($F.$lastSeen())
              \EndIf
              \State $Q.$enqueue($F'$)
          \Else
              \If {$v == F.$first()}
                 \comment{$v$ is the first $l$-mer of $F$, so $F$ is moved to the back of $Q$}
                 \State $Q$.requeue($F$)
              \ElsIf {sameOffSet($F.$first(), $v$)}
                 \State $F.$setLastSeen($v$)
              \EndIf
           \EndIf                             
       \EndIf
       \EndFor
       \State
       \comment{Final break up of families as needed; filter families with frequency $\leq f$.}
       \State tieLooseEnds()
  \EndFunction
\end{algorithmic}
\end{algorithm}

>>>>

We can now modify the previous definition of elementary repeats as
follows:
\begin{definition} Given a genomic sequence $G$, an integer $f$, and a
  spaced seed $s$, a sequence descriptor $r$ describes an elementary
  repeat if it meets the four (moified) requirements of an elementary repeat:
\begin{itemize}
\item \underline{Structure requriement}: $s$ covers $r$.
\item \underline{Frequency requirement}: There are at least $f$
  substrings of $G$ that match $r$.
\item \underline{Minimality requirement:} For every string $t$ in the
  decomposition of $r$ w.r.t $s$, the number of occurences of $t$ in
  $G$ is equal to the number of occurneces of $r$ in the genome.
\item \underline{Maximality requirement:} There is no sequence
  descriptor $r'$ of $r$ that contains $r$ as a proper substring and
  satisfies conditions 1-1.
\end{itemize}
\end{definition}

@@

We see in the attached table the pseudocode for the phRAIDER
algorithm, used to find a sequence descriptor for each set of
sequences that comprise an elementary repeat family with respoect to
some fixed seed $s$ and minimum frequency threshold $f$.  We see see
the core components of this algorithm as:
\begin{itemize}
\item

\begin{algorithm}
\renewcommand\thealgorithm{}
\caption{phRAIDER}
\begin{algorithmic}
  \Function{sameOffset}{$l$-mer $v_1$, $l$-mer $v_2$}
    \comment{Is the distance between the first two occurences of $v$ equal to the}
    \comment{the distance betwen the most recent two occurences?}
    \State \Return $H[v_2][0] - H[v_1][0]  == H[v_2][-1] - H[v_1][-1]$
  \EndFunction
  \State
  \Function{phRAIDER}{Genome $G$, Seed $s$, MinFrequence $f$}
    \comment{We are going to maintain a $Q$ of up to $|s|$ families, representing the families}
    \comment{of the last $|L|$ $l$-mers seen, sorted by the position of those last $l$-mers.}
    \State Queue$<$Family$>$ $Q$;
    \State
    \comment{Iterate over all $l$-mer positions in $G$}
    \For {$i \leftarrow 1 \to |G|-|s|$}

       \comment{Remove any families whose last $l$-mer did not occur in the last $i$ bases.}
       \While{$i - $location($Q$, finish, last, end)$ > |s|$}
         \State $Q.$dequeue()
       \EndWhile
       \State
       \comment{Get the $l$-mer occuring at $G_i$ and remove letters corresponding to $0$s in $s$.}
       \State $v \leftarrow \mbox{seeded}(G[i],x)$ 
       \State
       \comment{Add to list of $v$'s positions.}
       \State $H[v].push(i)$
       \State
       \If {$|H|(v)| == 2$}
          \comment{$v$ either is the beginning of a family, or should be combined with a}
          \comment{previous $l$-mer to form a family.}
          \State $F = \mbox{arg\_find}_{F \in Q}\{\mbox{sameOffset}(F.\mbox{start}(), v)\}$
          \If {$F$}
             \State $Q.$requeue($F$)
          \EndIf
          \State last($Q$).addLmer($v$)   
       \ElsIf {$|H[v]| > 2$}
          \comment{Is $v$ the proper continuation of a family, or does it need to be split off?}
          \State $F \leftarrow v.$family()
          \If {$H[F.\mbox{lastSeen}(i)][-1] < H[Q.\mbox{front}().\mbox{lastSeen}()]$}
              \comment{$v$ belongs to a family whose last instance is too far back.}
              \If {$v \neq F.$first$()$}
                  \comment{$v$ is not the first $l$-mer of its family}
                  \State $F' \leftarrow F.$split($F.v$)
              \Else
                  \State $F' \leftarrow F.$split($F.$lastSeen())
              \EndIf
              \State $Q.$enqueue($F'$)
          \Else
              \If {$v == F.$first()}
                 \comment{$v$ is the first $l$-mer of $F$, so $F$ is moved to the back of $Q$}
                 \State $Q$.requeue($F$)
              \ElsIf {sameOffSet($F.$first(), $v$)}
                 \State $F.$setLastSeen($v$)
              \EndIf
           \EndIf                             
       \EndIf
       \EndFor
       \State
       \comment{Final break up of families as needed; filter families with frequency $\leq f$.}
       \State tieLooseEnds()
  \EndFunction
\end{algorithmic}
\addtocounter{algorithm}{-1}
\end{algorithm}
